---
title: 'TidyVerse: ggplot2'
author: "Original from 'Kristin Lussi', I Souleymane Doumbia added my work below this student conclusion"
date: "2023-11-12"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

TidyVerse is a set of packages that share common data representations and API design. There are several packages in TidyVerse, such as ggplot2, dplyr, tidyr, readr, and several more. In this vignette, we will look at a few capabilities of the ggplot2 package using the student stress factors data from Kaggle.

### Load Packages

```{r, warning = FALSE, message = FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

### Load Data

We are working with a data set called "Student Stress Factors" from Kaggle. 

```{r, message = FALSE, warning = FALSE}
url <- "https://raw.githubusercontent.com/kristinlussi/FALL2023TIDYVERSE/main/Student%20Stress%20Factors.csv"

stress_factors <- read_csv(url, show_col_types = FALSE) %>%
  as.data.frame()

colnames(stress_factors) <- c("TimeStamp", "SleepQuality", "HeadacheFrequency", "AcademicPerformance", "StudyLoad", "ExtracurricularTimesPerWeek", "StressLevel")

head(stress_factors)
```

We will first need to create a data frame that will be easier to work with in ggplot2. 

```{r}
counts <- stress_factors %>%
  # pivot the factor columns into one column named "Type" with the values being the ranks
  pivot_longer(cols = c(SleepQuality, HeadacheFrequency, AcademicPerformance, StudyLoad, ExtracurricularTimesPerWeek, StressLevel), names_to = "Type", values_to = "Rank") %>%
  group_by(Type, Rank)
```

## Plots

### Bar Plot

Bar plots are useful for displaying categorical or discrete variables. In this section, we will demonstrate how to use ggplot2 to make a bar plot. Using the stress factor data, we will make a bar plot showing the counts of each rank grouped by category. 

The following code will produce a bar graph using ggplot2:

```{r}
# call ggplot
ggplot(data = counts, aes(x = Type, fill = Rank, group = Rank)) +
  # specify _bar graph
  geom_bar(position = "dodge") +
  # specify the title
  ggtitle("Rank Frequency for Each Stress Factor") +
  # specify the x label
  xlab("Stress Factor") +
  # specify the y label
  ylab("Count") +
  # center title
  theme(plot.title = element_text(hjust = 0.5),
        # title x categorical labels 45 degrees
        axis.text.x = element_text(angle = 45, hjust = 1)) 
```

In the above code, there are a few aesthetics called in the ggplot() function.

- "data" is set equal to "counts" to specify that we want to use the "counts" data frame
- aes() is short for "aesthetics". This function is used to define how variables in the dataset are mapped to visual properties of the plot
- "fill =" is set equal to the "Rank" variable. The "fill" aesthetic is used to assign colors or patterns to different groups or categories
- "group =" is set equal to the "Rank" variable. The "group" aesthetic is used to group data points by the "Rank" variable (e.g. to create separate dodge positions for each rank)

The following aesthetic is called in the geom_bar() function:

- "position =" is set equal to "dodge" to show the different groups side-by-side. There are several positions that can be called in ggplot such as stack, jitter, identity_dodge, etc...

### Violin Plot

In this section, we will demonstrate how to create a violin plot in ggplot2. Violin plots are useful when you want to observe the distribution of numeric data, especially whe you want to compare the distribution of different groups. 

The following code will produce a violin plot:

```{r}
ggplot(data = counts, aes(x = Type, y = Rank, fill = Type)) + 
  # specify that we want to create a violin plot
  geom_violin() +
  # specify title
  ggtitle("Rank Distribution for Each Stress Factor") +
  # specify x axis label
  xlab("Stress Factor") +
  # specify y axis label
  ylab("Rank") +
  # center title
  theme(plot.title = element_text(hjust = 0.5),
      # title x categorical labels 45 degrees
      axis.text.x = element_text(angle = 45, hjust = 1),
      # remove legend
      legend.position = "none") 
```

Similarly to the bar plot, the following aesthetics are called in the ggplot() function:

- "data" is set equal to "counts" data to specify that we want to use the "counts" data frame
- aes() is called and within the aes() function "x", "y", and "fill" are specified

### Density Plot

In this section, we will demonstrate how to create a density plot in ggplot2. Density plots are useful for visualizing the distribution of continuous data.

```{r}
ggplot(data = counts, aes(x = Rank, fill = Type)) +
  # specify that we want to create a density plot
  geom_density(alpha = 0.5) +
  # specify title 
  ggtitle("Distribution of Ranks for Each Stress Factor") +
  # specify x label
  xlab("Stress Factor") +
  # specify y label
  ylab("Density") +
  # center title
  theme(plot.title = element_text(hjust = 0.5)) 
```

The following aesthetics are called in the ggplot() function:

- "data" is set equal to "counts" to specify we want to use the "counts" data frame
- aes() contains "x = Rank" and "fill = Type" to specify that we want the different ranks on the x axis and we want different density lines for each stress factor

Within the geom_density() function, "alpha" is set equal to 0.5. The "alpha" aesthetic controls the transparency of the elements in the plot. 

## Plot Titles

In the previous plots, we can see that there are titles, x labels, and y labels. 

### Titles and Subtitles

To specify a title in ggplot2, we need to use the *ggtitle()* function.

To add a title and subtitle to a graph, for example, we would use the following code:

*ggtitle("Title", "Subtitle")*

To change positioning of the above items, we can use the *theme()* function with the following aesthetics:

- to center the title: *plot.title = element_text(hjust = 0.5)*
- to center the subtitle: *plot.subtitle = element_text(hjust = 0.5)*

### X and Y Labels

To specify the x-label in ggplot2, we need to use the *xlab()* function.

To specify the y-label in ggplot2, we need to use the *ylab()* function.

To change positioning of the above items, we can use the *theme()* function with the following aesthetics:

- to tilt the x-labels: *axis.text.x = element_text(angle = 45, hjust = 1)*

### Legends

When adding aesthetics to your plot, a legend will appear. To make changes to the legend, we can use the *theme()* function with the following aesthetics:

- to remove the legend: *legend.position = "none"*
- to add a title: *legend.title = " "*
- to change the legend position: *legend.position = " "*
- etc...

## Conclusion

In this brief document, we explored how to create several types of plots in ggplot2, including bar plots, violin plots, and density plots. ggplot2 offers a diverse range of functions for crafting data visualizations. Some of the most commonly used plot types include scatterplots, line plots, box plots, histograms, bar plots, and density plots.

## Sources

https://www.kaggle.com/datasets/samyakb/student-stress-factors

https://ggplot2.tidyverse.org 


# Souleymane Doumbia work start here:

## Additional Analysis: Exploring Relationships Between Variables"

### Exploring Relationships Between Variables
In addition to individual distributions, it's insightful to explore relationships between different stress factors. Here, we will create a scatter plot to visualize the relationship between 'Sleep Quality' and 'Stress Level'.

### Scatter Plot with ggplot2
A scatter plot is a useful tool to understand how two variables relate to each other. It can help in identifying trends, clusters, and outliers in the dataset.

```{r}
# Creating a scatter plot for Sleep Quality vs Stress Level
ggplot(data = stress_factors, aes(x = SleepQuality, y = StressLevel)) +
  geom_point(aes(color = StressLevel), alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Relationship Between Sleep Quality and Stress Level") +
  xlab("Sleep Quality") +
  ylab("Stress Level") +
  theme_minimal() +
  theme(legend.position = "none")
```
**In this plot:**
***aes(x = SleepQuality, y = StressLevel)*** sets the axes.
***geom_point()*** adds points to the plot, with their color indicating the stress level.
***geom_smooth(method = "lm", se = FALSE)*** adds a linear regression line to help identify trends.
***theme(legend.position = "none")*** removes the legend as it's redundant in this context.

###Interpretation###
The scatter plot allows us to observe if there is any apparent correlation between sleep quality and stress levels among students. The linear regression line gives a general trend, helping us to understand if better sleep quality is associated with lower stress levels, or vice versa.

###Final Conclusion###
This vignette has showcased the diverse capabilities of ggplot2 within the Tidyverse, using the "Student Stress Factors" dataset. We've seen how bar plots, violin plots, and density plots can visualize data distributions and frequencies. The addition of a scatter plot further enhances our analysis, revealing the relationship between sleep quality and stress levels among students.
Our exploration underscores ggplot2's versatility in data visualization, from simple plots to more complex relational analyses. This tool is invaluable for uncovering hidden trends and making informed conclusions from datasets. Through this journey, we've demonstrated how effective visualizations can transform raw data into meaningful insights, essential for data-driven decision-making.

* * *
