---
title: "marley_myrianthopoulos_tidyverse_extend"
author: "Marley Myrianthopoulos"
date: "`r Sys.Date()`"
output: html_document
---

## Summary

This is an extension of Michael Robinson's tidyverse assignment, which demonstrated how to use filter, arrange, group_by, and summarise. I will extend his work by demonstrating further use of the piping operator to perform multiple transforms simultaneously, and demonstrate that the results are the same.

```{r}
library(tidyverse)

patient_data <- read_csv("https://raw.githubusercontent.com/MRobinson112/tidy_verse/main/healthcare_dataset.csv")

```



## Piping Operator

The piping operator, "%>%" is used to take the output of one data transform and use it as the input of the next. This can be very useful if a number of transforms need to be procedurally applied. Michael demonstrates one method of multiple filtering below, and I will demonstrate another using multiple piping operators. An alternative (and less efficient) method is to create a new data frame for each transform.
```{r}

# Michael's method to filter all patients over 50 that are male 
male_patients_over_50_Michael <- patient_data %>%
  filter(Gender == "Male", Age > 50)

# My method to filter all patients over 50 that are male
male_patients_over_50_Marley <- patient_data %>%
  filter(Gender == "Male") %>%
  filter(Age > 50)

#Alternative Inefficient Method: This is what the piping operator is short-cutting.
male_patients <- patient_data %>%
  filter(Gender == "Male")
#Notice that the next line uses the male_patients data frame as its input, rather than patient_data
male_patients_over_50_alternative <- male_patients %>%
  filter(Age > 50)

#Order doesn't matter with multiple filters so we can reverse them
male_patients_over_50_Marley_2 <- patient_data %>%
  filter(Age > 50) %>%
  filter(Gender == "Male")

patients_over_50 <- patient_data %>%
  filter(Age > 50)
male_patients_over_50_alternative_2 <- patients_over_50 %>%
  filter(Gender == "Male")

# Observe that the outputs are identical for all of these different approaches

glimpse(male_patients_over_50_Michael)
glimpse(male_patients_over_50_Marley)
glimpse(male_patients_over_50_Marley_2)
glimpse(male_patients_over_50_alternative)
glimpse(male_patients_over_50_alternative_2)

```
